A sustainable approach to tourist signage on heritage trails

### 2023-11-22 paper.tex
aggiunto source combo_gen

Section \ref{sec:methods} provides the multidisciplinary framework. Section \ref{sec:legacy} illustrates the legacy emphasizing historical facts and locations. Section \ref{sec:tourism} describes the social and touristic context of the target location and provides the guidelines for integrating the initiative into the socio-political fabric. Such guidelines are used to motivate and define an agenda, in section \ref{sec:agenda}, and the section is concluded by a reasoned description of the area, highlighting the points of interest and their reachability \ref{sec:soa}. Section \ref{sec:implementation} describes the concrete implementation that follows the defined requirements. The description of the process to produce the printed tags starting from editable content on a geographic database allows reproduction and reuse of the solution in other contexts.

@@

A popular research direction covers the social aspects related to the use of smartphones. W. Tan \cite{tan17a} covers all aspects of a touristic experience related to the smartphone, from the definition of travel destination to assistance during the visit. Much attention is dedicated to the network of connections that is established thanks to the smartphone, which, again, is on the Internet. Such an assumption is in contrast with the title, which indicates a outdoor-based destination, where notably the Internet is not always reachable.

@@

In 2023 V. Rodrigues et al. published a systematic review of papers considering the interrelationships between tourism and portable digital devices \cite{rod23a}. Although the title evokes a one-way contribution, therefore assuming digitalization improves quality and sustainability of touristic offers, in the conclusions the authors reveal the awareness for the need to address {\em "the preservation of tourism attractions/sites"} and call for a {\em "a holistic approach ... to support a concept that still lacks conceptual and empirical clarification"}.

@@

Reading a tag does not require an ad-hoc application, and is performed approaching the smartphone to the tag: the user is advertised that content is available. The smartphone can read it aloud to compensate for the user's inability or translate to deal with linguistic issues. Such capabilities do not need an Internet connection but currently depend on ad-hoc applications.

@@

The QR code have been invented in 1994 for industrial applications. The current version, described in \cite{isoqr}, is dated February 2015.

In our use case it is especially appropriate when the cultural heritage extends across scarcely accessible land, considering that preserving and accessing sites located on private property or in hazardous areas can be challenging. In such cases, QR tags could improve the visitors’ experience by providing, when Internet coverage is available, the Web URLs for videos, aerial shots, and photos, even if the specific spot is not accessible. Such resources allow visitors to immerse themselves in the area's history and gain an experiential understanding.

@@

From a historical-geographical point of view, this area is identified with the \emph{Lima} stream and its tributaries, which impress the area with a peculiar geomorphology impervious despite the modest hilly elevations.

@@

The village of \emph{Casoli} is located south of the \emph{Lima} stream on a hill named {\em "Tanette"}. In the local small lair, the name reveals the presence of karstic cavities, some inhabited between the Paleolithic and the Iron Age and used also as stations on the trans-Apennine routes \cite{men76, gia96, pal63, zec72a, zec72b}.

@@

We have little evidence of Roman settlements in the mountainous hinterland, and the scarce archaeological traces are concentrated in the cave of \textit{Buca La Piella}, investigated in 1975 by the Centre for Archaeological Studies of \emph{Lucca}; it has two entrances joined by a walkable tunnel and rooms of discrete dimensions that overlook the outside. In addition to numerous faunal remains and fragments of locally produced common pottery, twenty bronze coins belonging to the 3rd century AD, bronze and lead objects were found \cite{gia96, men81, cia03}.

@@

As shown in \ref{tab:tourism}, the number of nights per tourist amounts to an average of 2,98 nights per tourist during a year and it is also significant since it shows, on one hand, \emph{Bagni di Lucca} context appeal for tourists choosing to stay more than a weekend and less than a week. On the other hand, this data means how long people likes staying in this tourist region and that they may have found attractions, services, and activities they need. Consequently, a well-organized touristic system attracts potentially touristic presences widespread on the municipality territory.

@@

From the point of view of our research, the context of \emph{Bagni di Lucca} helps to understand the complexity and variety of mixed tourism clusters (a sort of "touristic ecosystems" with common identitary elements such as: anthropic, socio-cultural, and historical features, as well as touristic services). Indeed, \emph{Val di Lima} can promote a very identifying destination brand straddling the town of \emph{Lucca} and the \emph{Garfagnana} mountain area. The Municipality of \emph{Bagni di Lucca} and the tourist area of the \emph{Val di Lima} are the context for the small and fascinating village of \emph{Casoli}, which is one of the 31 villages belonging to the Municipality. It is 7.22 km far from \emph{Bagni di Lucca}, at an altitude of 500 meters.

@@

According with the above guidelines for tourism valorization, we explain how the research has been conducted dealing with local stakeholders. Although information is still incomplete, we were able to collect sufficient opinions to confirm that \emph{Casoli} has the potential to be promoted through a community-involved tourism approach.

@@

In light of the above, a reflection on \textit{community-involved tourism} perspective is needed. As shown by this empirical study, a small village like \emph{Casoli} requires a valorization strategy which takes advantage of the wider local tourism context.

@@

\begin{figure}
\includegraphics[width =\linewidth]{figure/combo_gen.jpeg}
\caption{Generation of the QR tag and UMap visualization from UMap content. The uMap editor (bottom left) provides simplified access to the GeoJSON description. The same interface allows to specify the template controlling the Web visualization of the map (right side). An ad-hoc application referenced in the paper translates the uMap file downloaded from the server into a batch of printable QR tags (top left). The tag is highlighted in red in the picture on the Web display \label{fig:editing}}
\end{figure}

@@

Editing the map downloaded from the GPS app to fill in the {\em properties} does not require technical skills, since the tools integrated into the UMap user interface already fit the purpose (see the bottom left box in figure \ref{fig:editing}).

@@

The master document is rendered by UMap on the page linked to the tags as shown in figure \ref{fig:editing}.

@@

A rationale for the above attitude may be found in an article by Solima et al. \cite{sol18a}. The authors conducted a poll about user perception on two segments of users: Italians and Polish. The feelings about a QR based signage were not unanimous, and half of the Italian users considered it as not useful. Another result from the poll reveals that Polish users showed a stronger motivation to use tags for educational purposes (70\% versus 50\%), while Italians were mainly driven by curiosity. We conclude that, to provide a useful service, the signage must address a motivated user.

@@

To allow readability in {\em dead zones} and quickly provide the information without further steps to web pages, we encode the content within the QR code itself. The content consists of a text of an average of 460 characters aiming at a concise and effective communication with the tourist.

Therefore, in our design, the information encoded in the QR tag is the primary or only way of communication with the visitor. It is a new and original way of using such technology, since QR tags are instead considered ancillary to other forms of communication, like physical installations or linked Web contents, according with the motto "analog portal to the digital world", as in the title of \cite{bai12a}.

@@

One technology that fits the requirements of our task is Bluetooth Low Energy (BLE): the autonomy of a battery-operated BLE device is comparable with the periodic maintenance of signage based on passive devices. A Bluetooth device can send to the visitor's smartphone a beacon to indicate the presence of a passive sign, like a not-so-visible QR tag, or directly deliver information. However this requires the installation of an ad-hoc on the visitor's device, and the design of the IT product including the BLE device and the application, which may cost far more than the physical devices.

The demonstration of the existence of such needs biases the technological evolution: for instance, if counting NFC reads were a frequent requirement, tags with such a feature would be produced at low costs, and the increment operation might become a default in NFC reading apps. The many research results on low-energy and low-cost devices also indicates an expanding market.

@@

Tourism, like other productive activities, is at risk of being unsustainable, damaging the very resources that support it. To avoid heavy drawbacks the design of a tourism initiative has to consider the complex framework where it will operate according to an holistic approach. A methodology made of examples and guidelines helps to cope with such a challenging task.

@@

On a medium term perspective, the description of the deployed experiment gives suggestions for the development of new devices or features for existing ones. In our case, a new potential application may foster research aiming at passive or very low-energy devices that can effectively advertise their presence or simply collect statistics in \emph{dead-zones}.

@@

\bio
{\raisebox{-0.35cm}{\includegraphics[width=3.5cm,height=5.3cm,clip,keepaspectratio]{Definitions/augusto.png}}}
{\textbf{Augusto Ciuffoletti} Former teacher and researcher at the University of Pisa, retired on September 2023, with recent achievements on the Internet of Things and distributed control, with an interest on low-power devices. Previously active on Grid and, later, Cloud computing in collaboration with INFN. Until 2023, teaching activity at the Degree Course in Computer Humanities of the University of Pisa with two courses respectively on Network Protocols and the Angular 2+ framework.}
% For the MDPI journals use author-date citation, please follow the formatting guidelines on http://www.mdpi.com/authors/references
% To cite two works by the same author: \citeauthor{ref-journal-1a} (\citeyear{ref-journal-1a}, \citeyear{ref-journal-1b}). This produces: Whittaker (1967, 1975)
% To cite two works by the same author with specific pages: \citeauthor{ref-journal-3a} (\citeyear{ref-journal-3a}, p. 328; \citeyear{ref-journal-3b}, p.475). This produces: Wong (1999, p. 328; 2000, p. 475)

>>>>

Section \ref{sec:methods} provides the multidisciplinary framework. Section \ref{sec:legacy} illustrates the legacy emphasizing historical facts and locations. Section \ref{sec:tourism} describes the social and touristic context of the target location and provides the guidelines for integrating the initiative into the socio-political fabric. Such guidelines are used to motivate and define an agenda, in section \ref{sec:agenda}, and the section is concluded by a reasoned description of the area, highlighting the points of interest and their reachability \ref{sec:soa}. Section \ref{sec:implementation} describes the concrete implementation that follows the defined requirements. The description of the process to produce the printed tags starting from editable content on a geographic database allows the reproduction and reuse of the solution in other contexts.

@@

A popular research direction covers the social aspects related to the use of smartphones. W. Tan \cite{tan17a} covers all aspects of a touristic experience related to the smartphone, from the definition of travel destination to assistance during the visit. Much attention is dedicated to the network of connections that is established thanks to the smartphone, which, again, is on the Internet. Such an assumption is in contrast with the title, which indicates an outdoor-based destination, where notably the Internet is not always reachable.

@@

In 2023 V. Rodrigues et al. published a systematic review of papers considering the interrelationships between tourism and portable digital devices \cite{rod23a}. Although the title evokes a one-way contribution, therefore assuming digitalization improves the quality and sustainability of touristic offers, in the conclusions the authors reveal the awareness for the need to address {\em "the preservation of tourism attractions/sites"} and call for a {\em "a holistic approach ... to support a concept that still lacks conceptual and empirical clarification"}.

@@

Reading a tag does not require an ad-hoc application, and is performed by approaching the smartphone to the tag: the user is advertised that content is available. The smartphone can read it aloud to compensate for the user's inability or translate to deal with linguistic issues. Such capabilities do not need an Internet connection but currently depend on ad-hoc applications.

@@

The QR code was invented in 1994 for industrial applications. The current version, described in \cite{isoqr}, is dated February 2015.

In our use case, it is especially appropriate when the cultural heritage extends across scarcely accessible land, considering that preserving and accessing sites located on private property or in hazardous areas can be challenging. In such cases, QR tags could improve the visitors’ experience by providing, when Internet coverage is available, the Web URLs for videos, aerial shots, and photos, even if the specific spot is not accessible. Such resources allow visitors to immerse themselves in the area's history and gain an experiential understanding.

@@

From a historical-geographical point of view, this area is identified with the \emph{Lima} stream and its tributaries, which impress the area with peculiar geomorphology, impervious despite the modest hilly elevations.

@@

The village of \emph{Casoli} is located south of the \emph{Lima} stream on a hill named {\em "Tanette"}. In the local small lair, the name reveals the presence of karstic cavities, some inhabited between the Paleolithic and the Iron Age and also used as stations on the trans-Apennine routes \cite{men76, gia96, pal63, zec72a, zec72b}.

@@

We have little evidence of Roman settlements in the mountainous hinterland, and the scarce archaeological traces are concentrated in the cave of \textit{Buca La Piella}, investigated in 1975 by the Centre for Archaeological Studies of \emph{Lucca}; it has two entrances joined by a walkable tunnel and rooms of discrete dimensions that overlook the outside. In addition to numerous faunal remains and fragments of locally produced common pottery, twenty bronze coins from the 3rd century AD, bronze and lead objects were found \cite{gia96, men81, cia03}.

@@

As shown in \ref{tab:tourism}, the number of nights per tourist amounts to an average of 2,98 nights per tourist during a year and it is also significant since it shows, on one hand, \emph{Bagni di Lucca} context appeal for tourists choosing to stay more than a weekend and less than a week. On the other hand, this data means how long people like staying in this tourist region and that they may have found attractions, services, and activities they need. Consequently, a well-organized touristic system attracts potentially touristic presences widespread on the municipality territory.

@@

From the point of view of our research, the context of \emph{Bagni di Lucca} helps to understand the complexity and variety of mixed tourism clusters (a sort of "touristic ecosystems" with common identitary elements such as anthropic, socio-cultural, and historical features, as well as touristic services). Indeed, \emph{Val di Lima} can promote a very identifying destination brand straddling the town of \emph{Lucca} and the \emph{Garfagnana} mountain area. The Municipality of \emph{Bagni di Lucca} and the tourist area of the \emph{Val di Lima} are the context for the small and fascinating village of \emph{Casoli}, which is one of the 31 villages belonging to the Municipality. It is 7.22 km far from \emph{Bagni di Lucca}, at an altitude of 500 meters.

@@

According to the above guidelines for tourism valorization, we explain how the research has been conducted dealing with local stakeholders. Although information is still incomplete, we were able to collect sufficient opinions to confirm that \emph{Casoli} has the potential to be promoted through a community-involved tourism approach.

@@

In light of the above, a reflection on \textit{community-involved tourism} perspective is needed. As shown by this empirical study, a small village like \emph{Casoli} requires a valorization strategy that takes advantage of the wider local tourism context.

@@

\begin{figure}
\includegraphics[width =\linewidth]{figure/combo_gen.jpeg}
\caption{Generation of the QR tag and uMap visualization from uMap content. The uMap editor (bottom left) provides simplified access to the GeoJSON description. The same interface allows to specify the template controlling the Web visualization of the map (right side). An ad-hoc application referenced in the paper translates the uMap file downloaded from the server into a batch of printable QR tags (top left). The tag is highlighted in red in the picture on the Web display \label{fig:editing}}
\end{figure}

@@

Editing the map downloaded from the GPS app to fill in the {\em properties} does not require technical skills, since the tools integrated into the uMap user interface already fit the purpose (see the bottom left box in figure \ref{fig:editing}).

@@

The master document is rendered by uMap on the page linked to the tags as shown in figure \ref{fig:editing}.

@@

A rationale for the above attitude may be found in an article by Solima et al. \cite{sol18a}. The authors conducted a poll about user perception on two segments of users: Italians and Polish. The feelings about a QR-based signage were not unanimous, and half of the Italian users considered it as not useful. Another result from the poll reveals that Polish users showed a stronger motivation to use tags for educational purposes (70\% versus 50\%), while Italians were mainly driven by curiosity. We conclude that, to provide a useful service, the signage must address a motivated user.

@@

To allow readability in {\em dead zones} and quickly provide the information without further steps to web pages, we encode the content within the QR code itself. The content consists of a text of an average of 460 characters aiming at concise and effective communication with the tourist.

Therefore, in our design, the information encoded in the QR tag is the primary or only way of communication with the visitor. It is a new and original way of using such technology since QR tags are instead considered ancillary to other forms of communication, like physical installations or linked Web contents, according with to motto "analog portal to the digital world", as in the title of \cite{bai12a}.

@@

One technology that fits the requirements of our task is Bluetooth Low Energy (BLE): the autonomy of a battery-operated BLE device is comparable with the periodic maintenance of signage based on passive devices. A Bluetooth device can send to the visitor's smartphone a beacon to indicate the presence of a passive sign, like a not-so-visible QR tag, or directly deliver information. However, this requires the installation of an ad-hoc on the visitor's device, and the design of the IT product including the BLE device and the application, which may cost far more than the physical devices.

The demonstration of the existence of such needs biases the technological evolution: for instance, if counting NFC reads were a frequent requirement, tags with such a feature would be produced at low costs, and the increment operation might become a default in NFC reading apps. The many research results on low-energy and low-cost devices also indicate an expanding market.

@@

Tourism, like other productive activities, is at risk of being unsustainable, damaging the very resources that support it. To avoid heavy drawbacks the design of a tourism initiative has to consider the complex framework where it will operate according to a holistic approach. A methodology made of examples and guidelines helps to cope with such a challenging task.

@@

From a medium-term perspective, the description of the deployed experiment gives suggestions for the development of new devices or features for existing ones. In our case, a new potential application may foster research aiming at passive or very low-energy devices that can effectively advertise their presence or simply collect statistics in \emph{dead-zones}.

@@

\bio
{\raisebox{-0.35cm}{\includegraphics[width=3.5cm,height=5.3cm,clip,keepaspectratio]{Definitions/augusto.png}}}
{\textbf{Augusto Ciuffoletti} Former teacher and researcher at the University of Pisa, retired on September 2023, with recent achievements on the Internet of Things and distributed control, with an interest on low-power devices. Previously active on Grid and, later, Cloud computing in collaboration with INFN. Until 2023, he was a teacher at the Degree Course in Computer Humanities at the University of Pisa, where he taught two courses, one on Network Protocols and the other on the Angular 2+ framework.}
% For the MDPI journals use author-date citation, please follow the formatting guidelines on http://www.mdpi.com/authors/references
% To cite two works by the same author: \citeauthor{ref-journal-1a} (\citeyear{ref-journal-1a}, \citeyear{ref-journal-1b}). This produces: Whittaker (1967, 1975)
% To cite two works by the same author with specific pages: \citeauthor{ref-journal-3a} (\citeyear{ref-journal-3a}, p. 328; \citeyear{ref-journal-3b}, p.475). This produces: Wong (1999, p. 328; 2000, p. 475)
